---
title: "lab 17 part 2"
author: "Norman Lee A18086849"
format: html
---


>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. Hint: The read.table(),
summary() and boxplot() functions will likely be useful here. There is an example R script
online to be used ONLY if you are struggling in vein. Note that you can find the medium value
from saving the output of the boxplot() function to an R object and examining this object.
There is also the medium() and summary() function that you can use to check your
understanding

```{r}
expr <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
table(expr$geno)
nrow(expr)
tapply(expr$exp, expr$geno, median)
```
sizes:
A/A = 108
A/G = 233
G/G = 121

median    A/A      A/G      G/G 
31.24847 25.06486 20.07363 
```{r}
library(ggplot2)
```

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need
to be as polished as this one.

boxplot
```{r}
ggplot(expr)+aes(geno,exp, fill=geno)+ geom_boxplot()
```
A/A individuals have the highest expression, A/G individuals have medium expression, and G/G individuals have the lowest expression. Each additional G allele is associated with reduced expression, indicating that the G allele down-regulates.











