---
title: "hw7"
author: "Norman Lee"
date: "2025-11-03"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: sentence
---

---
title: "lab9hw"
author: "Norman Lee A18086849"
format: html
---

Download a CSV file from the PDB site (accessible from “Analyze” \> “PDB Statistics” \> “by Experimental Method and Molecular Type”. Move this CSV file into your RStudio project and use it to answer the following questions:

Q1: What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy.
**sum of x ray+ EM total/ sum of all values** \*\*(199931+29978)/244290\*100%= 94.5%\*\*

Q2: What proportion of structures in the PDB are protein?
\*\*239193/244290\*100%=97.9%\*\*

Q3: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?
**4866**

Q4: Water molecules normally have 3 atoms.
Why do we see just one atom per water molecule in this structure?
\***Each water molecule is represented by a single oxygen atom because hydrogens are invisible in X-ray structures**

Q5: There is a critical “conserved” water molecule in the binding site.
Can you identify this water molecule?
What residue number does this water molecule have **HOH 308**

Q6: Generate and save a figure clearly showing the two distinct chains of HIV-protease along with the ligand.
You might also consider showing the catalytic residues ASP 25 in each chain and the critical water (we recommend “Ball & Stick” for these side-chains).
Add this figure to your Quarto document.

![](images/1HSG%20(1).png)

```{r}
library(bio3d)
pdb <- read.pdb("1hsg")

pdb
```

Q7: How many amino acid residues are there in this pdb object?
198 Q8: Name one of the two non-protein residues?
HOH Q9: How many protein chains are in this structure?
2

```{r}
attributes(pdb)

```

Q10.
Which of the packages above is found only on BioConductor and not CRAN?
msa

Q11.
Which of the above packages is not found on BioConductor or CRAN?:
bio3dview

Q12.
True or False?
Functions from the pak package can be used to install packages from GitHub and BitBucket?
T

Q13.How many amino acids are in this sequence, i.e. how long is this sequence?
214 amino acids
